---
title: "SDMs of Leucobryum"
author: "Puwadol Chawengkul"
date: "8/8/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# 0. Set working environment

```{r Libraries}
library(dismo) #species distribution
library(raster) #read shape file
library(sp)
library(SDMtune)
library(tidyverse) #plot locations
library(maps) #map data
library(rasterVis)  #plot raster objects
library(geodata)
library(rJava)
library(zeallot)  #unpacking assignment (%<-%)

set.seed(1) #set random number for entire script
```

# 1. Environmental variables

## 1.1 Get bioclim and elevation data

Get the available bioclim data from the worldclim dataset. See more at [Bioclimatic variables](https://www.worldclim.org/data/bioclim.html)

```{r Get bioclim data}
bioclim.0 <- worldclim_country("Thailand", var = "bio", path = getwd(), res = 0.5)
bioclim <- stack(bioclim.0)
```

Get elevation from SRTM 90 m resolution data between -60 and 60 latitude

```{r Get elevation data}
elevation.0 <- elevation_30s("Thailand", path = getwd(), mask = TRUE, res = 0.5)
elevation.1 <- raster(elevation.0)
```

Make "elevation" be equal column to "bioclim" (for stacking)

```{r Stack preparation}
elevation <- resample(elevation.1, bioclim, method = "bilinear")
```

Stack predictors

```{r Stack predictors}
predictors <- stack(bioclim[[c(1:19)]], elevation) #choose bio1-19 and THA_elv

names(predictors)
```

## 1.2 Plot predictor

```{r bio 1}
gplot(predictors$wc2.1_30s_bio_19) +
    geom_tile(aes(fill = value)) +
    coord_equal() +
    scale_fill_gradientn(colours = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"),
                         na.value = "transparent",
                         name = "°C x 10") +
    labs(title = "Annual Mean Temperature",
         x = "longitude",
         y = "latitude") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
```

# 2. Species occurrence

## 2.1 Import from GBIF

```{r Import from GBIF}
gbif <- gbif("leucobryum","aduncum")
```

Clean data

```{r Clean GBIF data}
gbif.clean <- gbif %>% 
  #remove Leucobryum aduncum var. scalare
  filter(acceptedScientificName == "Leucobryum aduncum Dozy & Molkenboer, 1854") %>% 
  filter(fullCountry == "Thailand") %>% 
  select(lon, lat) %>% 
  na.omit()
```

## 2.2 Import from .csv

```{r Import from .csv}
csv <- read_csv("Leucobryum-Specimens.csv")
```

Clean data

```{r Clean .csv data}
csv.clean <- csv %>% 
  filter(Corrected_name == "Leucobryum aduncum") %>% 
  select(lon, lat) %>% 
  na.omit()
```

## 2.3 Merge GBIF and .csv

```{r Merge GBIF and .csv}
gbif.csv <- rbind(gbif.clean,csv.clean)
```

## 2.4 Species occurrence plot

```{r Species occurrence plot}
ggplot(map_data("world", region = "Thailand"), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
  geom_jitter(data = gbif.csv, aes(x = lon, y = lat), color = "red", alpha = 0.4, size = 1) +
  labs(x = "longitude", y = "latitude") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_fixed() +
  scale_x_continuous(limits = c(97, 106)) +
  scale_y_continuous(limits = c(5, 21))
```

## 2.5 Background Plot

```{r Create random background points}
bg <- randomPoints(predictors[[20]], 10000)
```

```{r Backgroun Plot}
ggplot(map_data("world", region = "Thailand"), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
  geom_jitter(data = as.data.frame(bg), aes(x = x, y = y), color = "red", alpha = 0.4, size = 1) +
  labs(x = "longitude", y = "latitude") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_fixed() +
  scale_x_continuous(limits = c(97, 106)) +
  scale_y_continuous(limits = c(5, 21))
```

## 2.6 SWD object

```{r Create SWD object}
data <- prepareSWD(species = "Leucobryum aduncum", p = gbif.csv, a = bg, env = predictors)

data
head(data@data)
head(data@coords)
data@species
```

```{r Save an SWD object}
#save a single file, pa indicating presence = 1 & background = 0
swd2csv(data, file_name = "data.csv")

#for two separate files
#swd2csv(data, file_name = c("presence.csv", "background.csv"))
```

```{r In Progress}
bg.cor.0 <- randomPoints(predictors, 10000)
bg.cor <- prepareSWD(species = "Bgs", a = bg.cor.0, env = predictors)

plotCor(bg.cor, method = "spearman", cor_th = 0.7)
corVar(bg.cor, method = "spearman", cor_th = 0.7)

c(train, test) %<-% trainValTest(data, test = 0.3, only_presence = TRUE, seed = 1)
maxnet_model <- train("Maxnet", data = train)

selected_variables_model <- varSel(maxnet_model, metric = "auc", test = test, bg4cor = bg.cor, method = "spearman", cor_th = 0.7, permut = 1)

map <- predict(selected_variables_model, data = predictors, type = "cloglog", file = "my_map", format = "GTiff")
```
