---
title: "SDMs of Leucobryum juniperoideum"
author: "Puwadol Chawengkul"
date: "27/04/2024"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# 0. Set working environment

```{r Libraries and extent}
library(dismo) #species distribution
library(raster) #read shape file
library(sf)
library(raptr) #random points
library(spThin) #occurrence thinning
library(SDMtune)
library(tidyverse) #plot locations
library(mapview) #plot locations on interactive map
library(maps) #map data
library(rasterVis)  #plot raster objects
library(geodata)
library(rJava)
library(zeallot)  #unpacking assignment (%<-%)
library(ENMeval)

set.seed(25) #set random number for entire script

#country <- gadm("Thailand", level = 0, path = getwd())
#country.ext <- as(country, "Spatial")
ext <- as(extent(88.59, 152.93, -12.38, 58.30), "SpatialPolygons")

#Fix bug, let R know the method
train(method = "Maxnet", data = data) #If you're running this code in the first time please skip and ignore error this line.
```

# 1. Environmental variables

## 1.1 Get bioclim

Get the available bioclim data from the worldclim dataset from [wc2.1_2.5m_bio.zip](https://biogeo.ucdavis.edu/data/worldclim/v2.1/base/wc2.1_2.5m_bio.zip), see Bioclimatic variables' code explanation at [Bioclimatic variables](https://www.worldclim.org/data/bioclim.html)

```{r Get bioclim data}
bio.files <- list.files("../WorldClim/wc2.1_2.5m_bio", ".tif$", full.names=T, recursive=T)
bioclim <- stack(bio.files)
names(bioclim) <- c("wc2_1","wc2_10","wc2_11","wc2_12","wc2_13","wc2_14","wc2_15","wc2_16","wc2_17","wc2_18","wc2_19","wc2_2","wc2_3","wc2_4","wc2_5","wc2_6","wc2_7","wc2_8","wc2_9")
```

Stack predictors

```{r Stack predictors}
predictors.0 <- stack(bioclim[[c(1:19)]]) #choose bio1-19
predictors.1 <- crop(predictors.0, ext)
predictors.2 <- mask(predictors.1, ext)
predictors <- terra::rast(predictors.2)
names(predictors)
```

## 1.2 Plot predictor

```{r bio 1}
#please check bio1 to bio19
gplot(predictors$wc2_1) +
    geom_tile(aes(fill = value)) +
    coord_equal() +
    scale_fill_gradientn(colours = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"),
                         na.value = "transparent",
                         name = "°C x 10") +
    labs(title = "Annual Mean Temperature",
         x = "longitude",
         y = "latitude") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
```

# 2. Species occurrence

## 2.1 Import from .csv

```{r Import from .csv}
csv <- read_csv("../Occurrence/data_20240422.csv")
```

Clean data

```{r Clean .csv data}
unique(csv$country)
unique(csv$specificEpithet)

occ.0 <- csv %>% 
  filter(specificEpithet == "juniperoideum") %>% 
  #filter(fullCountry == "Thailand") %>%
  select(genus, specificEpithet, longitude, latitude) %>% 
  na.omit()
```

## 2.2 Plot species occurrence

```{r Plot species occurrence with ggplot}
ggplot(map_data("world", region = "."), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
  geom_jitter(data = occ.0, aes(x = longitude, y = latitude), color = "red", alpha = 0.4, size = 1) +
  labs(x = "longitude", y = "latitude") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_fixed() + 
  scale_x_continuous(limits = c(88.59, 152.93)) +
  scale_y_continuous(limits = c(-12.38, 58.30))
```

## 2.3 Occurrence thinning

```{r Occurrence thinning}
thin(occ.0, lat.col = "latitude", long.col = "longitude",
     spec.col = "specificEpithet", thin.par = 4.5, reps = 1,
     out.dir = getwd())

occ <- read.csv("thinned_data_thin1.csv")
```

## 2.4 Plot thinned species occurrence

```{r Plot species occurrence with ggplot}
ggplot(map_data("world", region = "."), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
  geom_jitter(data = occ, aes(x = longitude, y = latitude), color = "red", alpha = 0.4, size = 1) +
  labs(x = "longitude", y = "latitude") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_fixed() + 
  scale_x_continuous(limits = c(88.59, 152.93)) +
  scale_y_continuous(limits = c(-12.38, 58.30))
```

## 2.5 Plot random background points

```{r Create random background points}
#bg.point <- raptr::randomPoints(predictors, 10000) # classic method 

occ.SpatialPoints <- SpatialPoints(occ[, c("longitude", "latitude")], proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))

buffer.0 <- buffer(occ.SpatialPoints, 500000) #circles with a radius of 5000 km
buffer <- aggregate(buffer.0)

bg.point <- spsample(buffer, n = 5000, type = "random")
```

```{r Plot random background points}
ggplot(map_data("world", region = "."), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
  geom_jitter(data = as.data.frame(bg.point), aes(x = x, y = y), color = "red", alpha = 0.4, size = 1) +
  labs(x = "longitude", y = "latitude") +
  theme(legend.position = "none") +
  coord_fixed() +
  scale_x_continuous(limits = c(88.59, 152.93)) +
  scale_y_continuous(limits = c(-12.38, 58.30))
```

# 3. Traning and evaluating the models

## 3.1 SWD object

```{r Create SWD object (species)}
data <- prepareSWD(species = "Leucobryum juniperoideum", p = occ[,2:3], a = bg.point, env = predictors)

data
head(data@data)
head(data@coords)
data@species
```

```{r Save an SWD object (species)}
#save a single file, pa indicating presence = 1 & background = 0
swd2csv(data, file_name = "data_swd_Leucobryum_juniperoideum.csv")

#for two separate files
#swd2csv(data, file_name = c("presence.csv", "background.csv"))
```

```{r Create SWD object (background)}
bg <- prepareSWD(species = "Bgs", a = bg.point, env = predictors)
```

```{r Explore variable correlation}
plotCor(bg, method = "spearman", cor_th = 0.7)
corVar(bg, method = "spearman", cor_th = 0.7)
```

```{r Train model with default setting}
#Basic train without cross validation
#c(train, test) %<-% trainValTest(data, test = 0.3, only_presence = TRUE, seed = 1)
#maxnet_model <- train("Maxnet", data = train)

#Fix varSel error message
data <- addSamplesToBg(data, all = TRUE)

#Spatial cross validation
block_folds <- get.block(occs = data@coords[data@pa == 1, ], bg = data@coords[data@pa == 0, ])
maxnet_model <- train(method = "Maxnet", data = data, fc = "lqph", reg = 1, folds = block_folds)

saveRDS(maxnet_model, file = "maxnet_model.RDS")
maxnet_model <- readRDS("maxnet_model.RDS")
maxnet_model

vi_maxnet_model <- varImp(maxnet_model, permut = 10)
vi_maxnet_model
plotVarImp(vi_maxnet_model)
cat("Training AUC: ", auc(maxnet_model))
cat("Testing AUC: ", auc(maxnet_model, test = TRUE))
plotROC(maxnet_model@models[[1]])
plotROC(maxnet_model@models[[2]])
plotROC(maxnet_model@models[[3]])
plotROC(maxnet_model@models[[4]])
```

```{r Remove highly correlated variables}
#jk <- doJk(maxnet_model, metric = "auc", test = test)
#jk
#plotJk(jk, type = "train", ref = auc(maxnet_model))
#plotJk(jk, type = "test", ref = auc(maxnet_model, test = test))

#Jackknife test has already included in VarSel()
selected_var_model <- varSel(maxnet_model, metric = "auc", bg4cor = bg, method = "spearman", cor_th = 0.7, permut = 10)

saveRDS(selected_var_model, file = "selected_var_model.RDS")
selected_var_model <- readRDS("selected_var_model.RDS")
selected_var_model

vi_selected_var_model <- varImp(selected_var_model, permut = 10)
vi_selected_var_model
plotVarImp(vi_selected_var_model)
cat("Training AUC: ", auc(selected_var_model))
cat("Testing AUC: ", auc(selected_var_model, test = TRUE))
plotROC(selected_var_model@models[[1]])
plotROC(selected_var_model@models[[2]])
plotROC(selected_var_model@models[[3]])
plotROC(selected_var_model@models[[4]])
```

```{r Tune model hyperparameters}
h <- list(reg = seq(0.5, 10, 0.5), fc = c("l", "lq", "lh", "lp", "lqp", "lqph", "lqpht"))
tune <- optimizeModel(selected_var_model, hypers = h, metric = "auc", gen = 20, seed = 25)
head(tune@results)
which.max(tune@results$test_AUC)
tune@models[[1]] #Best model

tuned_model <- train(method = "Maxnet", data = tune@models[[1]]@data, fc = tune@results[1, 1], reg = tune@results[1, 2], folds = block_folds)

saveRDS(tuned_model, file = "tuned_model.RDS")
tuned_model <- readRDS("tuned_model.RDS")
tuned_model

vi_tuned_model <- varImp(tuned_model, permut = 10)
vi_tuned_model
plotVarImp(vi_tuned_model)
write.csv(vi_tuned_model, "vi_tuned_model.csv", row.names = FALSE)
cat("Training AUC: ", auc(tuned_model))
cat("Testing AUC: ", auc(tuned_model, test = TRUE))
plotROC(tuned_model@models[[1]])
plotROC(tuned_model@models[[2]])
plotROC(tuned_model@models[[3]])
plotROC(tuned_model@models[[4]])
```

```{r Response curves}
plotResponse(tuned_model, var = "wc2_19", type = "cloglog", only_presence = TRUE, marginal = TRUE, fun = mean, rug = TRUE)
ggsave("response_curve_wc2_19.png", width = 7, height = 7, units = "cm")
ggsave("response_curve_wc2_19.pdf", width = 7, height = 7, units = "cm")

plotResponse(tuned_model, var = "wc2_7", type = "cloglog", only_presence = TRUE, marginal = TRUE, fun = mean, rug = TRUE)
ggsave("response_curve_wc2_7.png", width = 7, height = 7, units = "cm")
ggsave("response_curve_wc2_7.pdf", width = 7, height = 7, units = "cm")

plotResponse(tuned_model, var = "wc2_5", type = "cloglog", only_presence = TRUE, marginal = TRUE, fun = mean, rug = TRUE)
ggsave("response_curve_wc2_5.png", width = 7, height = 7, units = "cm")
ggsave("response_curve_wc2_5.pdf", width = 7, height = 7, units = "cm")
```

# 4. Making predictions

```{r Create a distribution map}
map.current.0 <- predict(tuned_model, data = predictors, type = "cloglog", filename = "map_current.tif", overwrite = TRUE)

map.current <- crop(map.current.0, ext)
```

```{r Plot a distribution map}
plotPred(map.current,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))

ggsave("map_current.png", width = 10, height = 7.5, units = "cm")
ggsave("map_current.pdf", width = 10, height = 7.5, units = "cm")
```

# 5. Future projections

## 5.1 Get future bioclim ssp 245

```{r Future ssp 245 model 1}
cmip6.45.1.0 <- cmip6_world("CNRM-CM6-1", "245", "2061-2080", var = "bioc", res = 2.5, path = "../WorldClim/")
cmip6.45.1.1 <- crop(cmip6.45.1.0, ext)
cmip6.45.1.2 <- stack(cmip6.45.1.1[[c(1:19)]]) #choose bio1-19
cmip6.45.1 <- terra::rast(cmip6.45.1.2)
names(cmip6.45.1)

gplot(cmip6.45.1$wc2_1) +
    geom_tile(aes(fill = value)) +
    coord_equal() +
    scale_fill_gradientn(colours = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"),
                         na.value = "transparent",
                         name = "°C x 10") +
    labs(title = "Annual Mean Temperature",
         x = "longitude",
         y = "latitude") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
```

```{r Future ssp 245 model 2}
cmip6.45.2.0 <- cmip6_world("EC-Earth3-Veg", "245", "2061-2080", var = "bioc", res = 2.5, path = "../WorldClim/")
cmip6.45.2.1 <- crop(cmip6.45.2.0, ext)
cmip6.45.2.2 <- stack(cmip6.45.2.1[[c(1:19)]]) #choose bio1-19
cmip6.45.2 <- terra::rast(cmip6.45.2.2)
names(cmip6.45.2)
```

```{r Future ssp 245 model 3}
cmip6.45.3.0 <- cmip6_world("MPI-ESM1-2-HR", "245", "2061-2080", var = "bioc", res = 2.5, path = "../WorldClim/")

cmip6.45.3.1 <- crop(cmip6.45.3.0, ext)
cmip6.45.3.2 <- stack(cmip6.45.3.1[[c(1:19)]]) #choose bio1-19
cmip6.45.3 <- terra::rast(cmip6.45.3.2)
names(cmip6.45.3)
```

```{r Future ssp 245 model 4}
cmip6.45.4.0 <- cmip6_world("MPI-ESM1-2-LR", "245", "2061-2080", var = "bioc", res = 2.5, path = "../WorldClim/")
cmip6.45.4.1 <- crop(cmip6.45.4.0, ext)
cmip6.45.4.2 <- stack(cmip6.45.4.1[[c(1:19)]]) #choose bio1-19
cmip6.45.4 <- terra::rast(cmip6.45.4.2) 
names(cmip6.45.4)
```

```{r Future ssp 245 model 5}
cmip6.45.5.0 <- cmip6_world("MRI-ESM2-0", "245", "2061-2080", var = "bioc", res = 2.5, path = "../WorldClim/")
cmip6.45.5.1 <- crop(cmip6.45.5.0, ext)
cmip6.45.5.2 <- stack(cmip6.45.5.1[[c(1:19)]]) #choose bio1-19
cmip6.45.5 <- terra::rast(cmip6.45.5.2)
names(cmip6.45.5)
```

```{r Future ssp 245 model 6}
cmip6.45.6.0 <- cmip6_world("UKESM1-0-LL", "245", "2061-2080", var = "bioc", res = 2.5, path = "../WorldClim/")
cmip6.45.6.1 <- crop(cmip6.45.6.0, ext)
cmip6.45.6.2 <- stack(cmip6.45.6.1[[c(1:19)]]) #choose bio1-19
cmip6.45.6 <- terra::rast(cmip6.45.6.2)
names(cmip6.45.6)
```

## 5.2 Making future predictions

```{r Create the future distribution maps}
map.45.1.0 <- predict(tuned_model, data = cmip6.45.1, type = "cloglog", file = "map_future_1.tif", overwrite = TRUE)
map.45.1 <- crop(map.45.1.0, ext)

map.45.2.0 <- predict(tuned_model, data = cmip6.45.2, type = "cloglog", file = "map_future_2.tif", overwrite = TRUE)
map.45.2 <- crop(map.45.2.0, ext)

map.45.3.0 <- predict(tuned_model, data = cmip6.45.3, type = "cloglog", file = "map_future_3.tif", overwrite = TRUE)
map.45.3 <- crop(map.45.3.0, ext)

map.45.4.0 <- predict(tuned_model, data = cmip6.45.4, type = "cloglog", file = "map_future_4.tif", overwrite = TRUE)
map.45.4 <- crop(map.45.4.0, ext)

map.45.5.0 <- predict(tuned_model, data = cmip6.45.5, type = "cloglog", file = "map_future_5.tif", overwrite = TRUE)
map.45.5 <- crop(map.45.5.0, ext)

map.45.6.0 <- predict(tuned_model, data = cmip6.45.6, type = "cloglog", file = "map_future_6.tif", overwrite = TRUE)
map.45.6 <- crop(map.45.6.0, ext)
```

```{r Plot the future distribution maps}
#Current distribution map
plotPred(map.current,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_current.png", width = 10, height = 7.5, units = "cm")
ggsave("map_current.pdf", width = 10, height = 7.5, units = "cm")

#Future distribution map 1
plotPred(map.45.1,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_future_1.png", width = 10, height = 7.5, units = "cm")
ggsave("map_future_1.pdf", width = 10, height = 7.5, units = "cm")

#Future distribution map 2
plotPred(map.45.2,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_future_2.png", width = 10, height = 7.5, units = "cm")
ggsave("map_future_2.pdf", width = 10, height = 7.5, units = "cm")

#Future distribution map 3
plotPred(map.45.3,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_future_3.png", width = 10, height = 7.5, units = "cm")
ggsave("map_future_3.pdf", width = 10, height = 7.5, units = "cm")

#Future distribution map 4
plotPred(map.45.4,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_future_4.png", width = 10, height = 7.5, units = "cm")
ggsave("map_future_4.pdf", width = 10, height = 7.5, units = "cm")

#Future distribution map 5
plotPred(map.45.5,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_future_5.png", width = 10, height = 7.5, units = "cm")
ggsave("map_future_5.pdf", width = 10, height = 7.5, units = "cm")

#Future distribution map 6
plotPred(map.45.6,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_future_6.png", width = 10, height = 7.5, units = "cm")
ggsave("map_future_6.pdf", width = 10, height = 7.5, units = "cm")

#Mean future distribution map
future.mean <- mean(map.45.1,map.45.2,map.45.3,map.45.4,map.45.5,map.45.6)

plotPred(future.mean,
         lt = "Habitat\nsuitability",
         colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
ggsave("map_future_mean.png", width = 10, height = 7.5, units = "cm")
ggsave("map_future_mean.pdf", width = 10, height = 7.5, units = "cm")
```

```{r Plot the presence/absence maps}
ths.1 <- thresholds(tuned_model@models[[1]], type = "cloglog")
ths.2 <- thresholds(tuned_model@models[[2]], type = "cloglog")
ths.3 <- thresholds(tuned_model@models[[3]], type = "cloglog")
ths.4 <- thresholds(tuned_model@models[[4]], type = "cloglog")
ths <- mean(c(ths.1[3, 2], ths.2[3, 2], ths.3[3, 2], ths.4[3, 2]))
ths

#Current presence/absence map
plotPA(map.current, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_current.tif", overwrite = TRUE)
ggsave("pa_map_current.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_current.pdf", width = 10, height = 7.5, units = "cm")

#Future presence/absence maps
plotPA(map.45.1, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_future_1.tif", overwrite = TRUE)
ggsave("pa_map_future_1.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_1.pdf", width = 10, height = 7.5, units = "cm")
plotPA(map.45.2, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_future_2.tif", overwrite = TRUE)
ggsave("pa_map_future_2.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_2.pdf", width = 10, height = 7.5, units = "cm")
plotPA(map.45.3, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_future_3.tif", overwrite = TRUE)
ggsave("pa_map_future_3.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_3.pdf", width = 10, height = 7.5, units = "cm")
plotPA(map.45.4, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_future_4.tif", overwrite = TRUE)
ggsave("pa_map_future_4.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_4.pdf", width = 10, height = 7.5, units = "cm")
plotPA(map.45.5, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_future_5.tif", overwrite = TRUE)
ggsave("pa_map_future_5.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_5.pdf", width = 10, height = 7.5, units = "cm")
plotPA(map.45.6, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_future_6.tif", overwrite = TRUE)
ggsave("pa_map_future_6.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_6.pdf", width = 10, height = 7.5, units = "cm")

plotPA(future.mean, th = ths, colors = c("#018571", "#d8b365"), filename = "pa_map_future_mean.tif", overwrite = TRUE)
ggsave("pa_map_future_mean.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_mean.pdf", width = 10, height = 7.5, units = "cm")
```

```{r Consensus of the future presence/absence maps}
future.maps <- stack("pa_map_future_1.tif", "pa_map_future_2.tif", "pa_map_future_3.tif", "pa_map_future_4.tif", "pa_map_future_5.tif", "pa_map_future_6.tif")

future.sum <- sum(future.maps)
gplot(future.sum) +
    geom_tile(aes(fill = value)) +
    coord_equal() +
    scale_fill_gradientn(colours = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"),
                         na.value = "transparent",
                         name = "Frequency") +
    labs(title = "The Consensus Map of the Future",
         x = "Longitude (°)",
         y = "Latitude (°)") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
ggsave("pa_map_future_sum.png", width = 10, height = 7.5, units = "cm")
ggsave("pa_map_future_sum.pdf", width = 10, height = 7.5, units = "cm")
```

```{r Consensus of the future presence/absence}
#Get sizes of all cells in raster (km2)
cell_size <- cellSize(future.mean, mask = TRUE, unit = "km")
#Delete NAs from vector of all raster cells
cell_size <- cell_size[!is.na(cell_size)]
#Compute area (km2) of all cells
raster_area <- length(cell_size)*median(cell_size)
#Print area according to raster object
cat("Study area:", raster_area, "km2")

future_freq_consensus <- as.data.frame(freq(future.sum, useNA = "no")) %>% 
  mutate(area_km2 = count/sum(count)*raster_area) %>% 
  mutate(percentage = round(count/sum(count)*100, 2))
future_freq_consensus

write.csv(future_freq_consensus, "future_freq_consensus.csv", row.names = FALSE)
```

# 6. Future changing distribution

## 6.1 Plot the changing distribution map

```{r Calculate the diffence between current and future}
pa_map_current <- raster("pa_map_current.tif")
pa_map_future <- raster("pa_map_future_mean.tif")
pa_map_change <- (2*pa_map_future) - pa_map_current
#Values: increase area = 2, consistent area = 1, decrease area = -1,
#        while 0 represented absent area in both current and future scenarios

pa_map_change_df.0 <- as.data.frame(pa_map_change, xy = TRUE)
pa_map_change_df <- pa_map_change_df.0 %>% 
  mutate(layer.2 = cut(layer, breaks = c(-1.5, -0.5, 0.5, 1.5, 2.5)))
```

```{r Plot the changing distribution map}
ggplot() +
  geom_raster(data = pa_map_change_df, 
              aes(x = x, y = y, fill = layer.2)) +
  labs(x = "Longitude (°)", y = "Latitude (°)") +
  scale_fill_manual(values = c("tomato", "grey", "cornflowerblue", "green"), 
                    name = "Class",
                    labels = c("Disappearance", "Continued absence", "Continued presence", "Gaining", "Sea"),
                    na.value = "NA") + 
  geom_jitter(data = occ, aes(x = longitude, y = latitude), color = "black", alpha = 0.4, size = 1) +
  scale_x_continuous(limits = c(88.59, 152.93)) +
  scale_y_continuous(limits = c(-12.38, 58.30)) +
  coord_quickmap() +
  theme_minimal()
ggsave("pa_map_change.png", width = 15, height = 11.25, units = "cm")
ggsave("pa_map_change.pdf", width = 15, height = 11.25, units = "cm")
```

## 6.2 Summary of the area

```{r Current present area}
pa_current_freq <- as.data.frame(freq(pa_map_current, useNA = "no")) %>% 
  mutate(meaning = c("absent", "present")) %>% 
  mutate(area_km2 = count/sum(count)*raster_area) %>% 
  mutate(percentage = round(count/sum(count)*100, 2))

pa_current_freq
write.csv(pa_current_freq, "pa_current_freq.csv", row.names = FALSE)

cat("Current present area:", pa_current_freq$area_km2[2], "km2")
```

```{r Future present area}
pa_future_freq <- as.data.frame(freq(pa_map_future, useNA = "no")) %>% 
  mutate(meaning = c("absent", "present")) %>% 
  mutate(area_km2 = count/sum(count)*raster_area) %>% 
  mutate(percentage = round(count/sum(count)*100, 2))

pa_future_freq
write.csv(pa_future_freq, "pa_future_freq.csv", row.names = FALSE)

cat("Future present area:", pa_future_freq$area_km2[2], "km2")
```

```{r Future present area}
pa_change_freq <- as.data.frame(freq(pa_map_change, useNA = "no")) %>% 
  mutate(meaning = c("decrease", "absent", "consistent", "increase")) %>% 
  mutate(area_km2 = count/sum(count)*raster_area) %>% 
  mutate(percentage = round(count/sum(count)*100, 2)) %>% 
  mutate(percentage_diff_current_present = round(area_km2/pa_current_freq$area_km2[2]*100, 2))

pa_change_freq
write.csv(pa_change_freq, "pa_change_freq.csv", row.names = FALSE)
```
